---
title: "Correlation climate-poverty"
author: "Jacques Sauve"
date: "21 de dezembro de 2015"
output: html_document
---

**This file not to be considered**.
This is a scratch file with some old code and observations.
**This file not to be considered**.


TODO
- to be visually attractive, show relationships, not prediction. do a regression with up to 10 inputs
- Things to try:
    + plan interface
        + inputs: region, log?, box-cox?, gdp/poverty, predictors to use
        + outputs:
            + if 1 predictor: scatter plot, R^2
            + else: standardized slope, partial regressions, R^2, 
            % explained by each predictor
            + 3d plot if 2 predictors
    + center and scale, box-cox
    + include region and finish a first shiny app
    + Pre-process data (log GDP, box-cox, or of other metric used)
    + Better to use the temperature of the capital city
    + Use adjusted PPP and constant dollars
    + Use another index of poverty instead of gdp
    + include other factors that man cannot change (eg. natural resources)
    + Find clusters (Europe, ...)

- is there a correlation between climate and poverty. choropleth, regression models, factors other than climate?
- do the same inside europe, or countries like Italy, Brazil, ...
- include idh (as predictor? as response?)
- evolution over time
- interactive shiny app
- executive summary, how to use app
- bokeh

# Read data

## Read in economic data.

```{r}
df.econ = read.csv('Popular indicators/Popular indicators_Data.csv')
head(df.econ)
unique(df.econ$Series.Name)

# average over all years
ss = subset(x = df.econ,
            select = grep('^X', colnames(df.econ)))
df.econ$mean = rowMeans(ss, na.rm = TRUE)

gdp.per.capita = subset(x = df.econ, 
                    subset = (Series.Name == 'GDP per capita (current US$)'))
names(gdp.per.capita)[names(gdp.per.capita) == 'mean'] <- 'mean.gdp'
dim(gdp.per.capita)
head(gdp.per.capita)

poverty = subset(x = df.econ,
                 subset = (Series.Name == 'Poverty headcount ratio at national poverty lines (% of population)'))
names(poverty)[names(poverty) == 'mean'] <- 'mean.poverty'
dim(poverty)
head(poverty)
```

## Read in temperature data.

```{r message=FALSE, warning=FALSE}
library(gdata)
temp.df = read.xls(xls = 'cckp_historical_data_0.xls', sheet = 4, na.strings = 'NA')
head(temp.df)
dim(temp.df)
```

## Read in natural resources data.

Natural resource rent is not completely independent of man becsuse it indicates that the natural resources are being exploited.
Still, it seems to be a good proxy for natural resources that a country has.

```{r}
nr.df = read.csv('ny.gdp.totl.rt.zs_Indicator_en_csv_v2/ny.gdp.totl.rt.zs_Indicator_en_csv_v2.csv', skip=4)
dim(nr.df)
head(nr.df)
unique(nr.df$Indicator.Name)
# average over all years
ss = subset(x = nr.df,
            select = grep('^X', colnames(nr.df)))
nr.df$mean.nr = rowMeans(ss, na.rm = TRUE)
head(nr.df)
```

## Create main data frame

```{r}
comb.df = merge(gdp.per.capita, temp.df, 
                by.x='Country.Code', by.y='ISO_3DIGIT', 
                all=FALSE, sort=TRUE)
comb.df = merge(comb.df, poverty, 
                by.x='Country.Code', by.y='Country.Code', 
                all=FALSE, sort=TRUE)
comb.df = merge(comb.df, nr.df, 
                by.x='Country.Code', by.y='Country.Code', 
                all=FALSE, sort=TRUE)
df = subset(comb.df, select=c('Country.Code', 'Annual_temp', 'mean.gdp', 'mean.poverty', 'mean.nr'))
head(df)
dim(df)
```
 
# Regressions

- Use log of GDP, since many orders of magnitude are involved.

```{r warning=FALSE}
library(ggplot2)
set.seed(1)
cor(df$Annual_temp, log(df$mean.gdp), use='pairwise.complete.obs')
g = ggplot(df, aes(Annual_temp, log(mean.gdp))) +
    geom_point() +
    geom_text(aes(label=Country.Code)) +
    geom_smooth(method=lm)
g
fit = lm(log(mean.gdp) ~ Annual_temp , data=df)
coef(fit)
summary(fit)
```

- There is a correlation (-0.46) between log GDP per capita and temperature.
    + Only about 20% of the variation is explained by temperature
    I expected a lot more!
    + Oil countries seem to have high GDP but maybe not have a rich population
        + Maybe we need another response that better reflects how rich the average population is
    + We clearly need more factors to explain GDP

- Try another metric: poverty index

```{r warning=FALSE}
set.seed(1)
cor(df$Annual_temp, df$mean.poverty, use='pairwise.complete.obs')
g = ggplot(df, aes(Annual_temp, mean.poverty)) +
    geom_point() +
    geom_text(aes(label=Country.Code)) +
    geom_smooth(method=lm)
g
fit = lm(log(mean.poverty) ~ Annual_temp , data=df)
coef(fit)
summary(fit)
```

- many countries are missing
- temperature explains 20% of poverty
- Essentially no difference in results between using gdp and poverty index as response

## Regression with two predictors: temperature and natural resources

```{r}
set.seed(1)
fit = lm(log(mean.gdp) ~ Annual_temp + mean.nr, data=df)
coef(fit)
summary(fit)
```

- Explains 21%, just a bit more
- Try absolute natural resource rent, not percentage of GDP

```{r}
set.seed(1)
df$mean.nr.abs = df$mean.nr * df$mean.gdp
fit1 = lm(log(mean.gdp) ~ Annual_temp + log(mean.nr.abs), data=df)
coef(fit1)
summary(fit1)
```

- We're now explaining 39%, a lot more!

## Extracting some additional metrics from the data

```{r}
# partition the R^2 in the various predictors
# (careful: this may depend on the order of predictors.
# It only really works if the predictors are orthogonal)
percent.explained = function(model) {
    an <- anova(model)
    anss <- an$"Sum Sq"
    return (cbind(an,PctExp=anss/sum(anss)*100))
}
print(percent.explained(fit1))
```

- Temperature explains 22% of the variation
- Natural resources explain an additional 18%
- We still have 60% unexplained
- But for this little exercise, 2 predictors is enough

- Let's redo the last model with normalized variables

```{r}
library(caret)
set.seed(1)
fit2 = train(log(mean.gdp) ~ Annual_temp + log(mean.nr.abs), 
             data=df, method='lm',
             preProcess = c('center', 'scale'))
print(percent.explained(fit2$finalModel))
```

- Nothing really changes

```{r}
set.seed(1)
fit3 = train(mean.gdp ~ Annual_temp + mean.nr.abs, 
             data=df, method='lm',
             preProcess = c('center', 'scale'))
print(percent.explained(fit3$finalModel))
```

- No real change

# Some plots

```{r}
library(car)
library(ggplot2)
# temperature plot, given the natural resources

g = ggplot(df, aes(Annual_temp, log(mean.gdp))) +
    geom_point() +
    geom_text(aes(label=Country.Code)) +
    geom_smooth(method=lm)
g

# Added-variable plots: regression using one variable, maintaining others constant
avPlots(fit2$finalModel, main='Partial Regression Plots', labels=df[complete.cases(df),]$Country.Code, id.n=20, layout=c(2,1))

# plot of standardized coefficients with 2 SD confidence intervals
library(arm)
coefplot(fit2$finalModel, varnames=c('Intercept', 'Average\nAnnual\nTemperature', 'log\nNatural\nResources'), cex.var=1, cex.pts=2, main='Standardized coefficients')

# 3d plot
library(scatterplot3d)
s3d <- scatterplot3d(x=df$Annual_temp, y=log(df$mean.nr.abs), z=log(df$mean.gdp), type = "h", color = "blue",
angle = 55, scale.y = 0.7, pch = 16, main = "Regression of GDP")
s3d$plane3d(fit2$finalModel)
```

- 3D plot hard to see: doesn't add anything

# References, Data Sources

- GDP per capita
    + http://databank.worldbank.org/data/reports.aspx?Code=NY.GDP.PCAP.CD&id=af3ce82b&report_name=Popular_indicators&populartype=series&ispopular=y#
    + advanced options: NA preference = NA
- Temperature data
    + http://data.worldbank.org/data-catalog/cckp_historical_data
- Natural Resource Rent data
    + http://data.worldbank.org/indicator/NY.GDP.TOTL.RT.ZS
- [How to attribute variance explained among predictors](http://stats.stackexchange.com/questions/79399/calculate-variance-explained-by-each-predictor-in-multiple-regression-using-r)
- [How to visualize multiple regression](http://stats.stackexchange.com/questions/89747/how-to-describe-or-visualize-a-multiple-linear-regression-model)
- [Country geo data](http://mledoze.github.io/countries/)

pe = percent.explained(fit$finalModel)
ggplot(pe, aes(x=row.names(pe), y=PctExp)) +
geom_bar(stat='identity', fill='blue') + 
geom_text(aes(x=row.names(pe), y=PctExp+0.8, label=format(PctExp, digits=2)))
xlab('Predictor') +
ylab('Percent Variation Explained')